A user-extensible and adaptable parser architecture

### 2008-08-12 doc/sgai-2008.tex
Dealt with some XXX items: * Mentioned prior art. * Added a little more about detecting overlapping regexes. * Removed "Why are there dips in the graphs?" paragraphs, they weren't   particularly important any more. * Changed normal to optimal.

% \showgraph{filename}{caption}{label}
\newcommand{\showgraph}[3]{%
    \begin{figure}[hbt!]%
        \caption{#2}\label{#3}%
        \includegraphics{#1}%
    \end{figure}%
}

%\showtable{filename}{caption}{label}
\newcommand{\showtable}[3]{
    \begin{table}[ht]
        \caption{#2}\label{#3}
        \input{#1}
    \end{table}
}

@@

XXX DISCUSS COMPLEXITY\@?  EVEN JUST MENTION THAT COMPLEXITY RISES WHEN
INCREMENTAL PARSING IS USED, BUT THAT EXTERNAL TOOLS COULD BE USED TO DEAL
WITH IT, AND THAT IT IS ENTIRELY DEPENDANT ON THE RULESET USED\@?

XXX MAYBE MERGE THE AUTHOR SECTIONS IF DESPERATE FOR SPACE\@.

XXX REVIEWER SUGGESTED COMPARING TO EXISTING WORK, BUT I DID NOT FIND ANY
SIMILAR EXISTING WORK\@.  I SHOULD REALLY SEARCH AGAIN\@.

XXX OVERLAP DETECTION\@: L1 n L2 = E* --- ((E* --- L1) u (E* --- L2)).
PERL REGEXES ARE AT LEAST AS POWERFUL AS A PUSHDOWN AUTOMATA --- SEE
CHOMSKY'S HIERARCHY TO MAKE SURE\@.  CONTAINMENT CANNOT BE DETERMINED FOR
CFGs, INTERSECTION IS POSSIBLE\@.

@@

Existing parsers were considered, but their parsing coverage was too low,
their parsing too inexact, and/or they did not extract sufficient data.
The effort required to adapt and improve an existing parser was judged to
be greater than the effort to write one from scratch, as the techniques
used by the existing parsers severely limited their potential: some ignored
the majority of log lines, parsing specific log lines with great accuracy,
but in an inextensible fashion; others sloppily parsed the majority of log
lines, but were incapable of distinguishing between log lines of the same
type, e.g.\ rejections, and extracted only a portion of the data required.

@@

The parser does not try to detect overlapping rules; that responsibility is
left to the author of the rules.  Unintentionally overlapping rules lead to
inconsistent parsing and data extraction because the first matching rule
wins, and the order in which rules are tried against each line may change
between log files.  Overlapping rules are frequently a requirement,
allowing a more specific rule to match some log lines and a more general
rule to match the majority, e.g.\ separating SMTP delivery to specific
sites from SMTP delivery to the rest of the world.  The algorithm provides
a facility for specifying the order of overlapping rules: the priority
field in each rule (defaults to zero).  Rules are sorted by priority,
highest first, and then rules with the same priority are sorted by the
number of successful matches they had parsing the previous log file.
Negative priorities may be useful for catchall rules.  XXX DETECTING
OVERLAP\@.

@@

    \item [Partially context-sensitive actions and cascaded parsing:]
        Actions can consult the results (or lack of results) of previous
        actions when executing, providing some context sensitivity.
        Actions can not inspect past or upcoming inputs, however actions
        can return a modified input line that will be parsed as if read
        from the input stream, allowing for a simplified version of
        cascaded parsing~\cite{cascaded-parsing}.

@@

An important property of a parser is how parsing time scales relative to
input size: does it scale linearly, polynomially, or exponentially?
\Graphref{Parsing time, file size, and number of lines} shows the parsing
time in seconds, file size in MB, and tens of thousands of log lines for
each of the \numberOFlogFILES{} log files.  All three lines run roughly in
parallel, giving the impression that the algorithm scales linearly with
input size.  This impression is borne out by \graphref{parsing time vs file
size vs number lines factor} which plots both the ratio of file size vs
parsing time, and the ratio of number of log lines vs parsing time (higher
is better in both cases).  The ratios are quite tightly banded across the
graph, showing that the algorithm scales linearly.  The ratio increases
(i.e.\ improves) for log files 22 and 62--68 despite their larger than
average size (shown in \graphref{Parsing time, file size, and number of
lines}); this is explained in \sectionref{Why are there dips in the
graphs?}.  \showgraph{build/graph-input-size-vs-parsing-time}{Parsing time,
file size, and number of lines}{Parsing time, file size, and number of
lines} \showgraph{build/graph-input-size-vs-parsing-time-ratio}{Ratio of
file size and number of lines to parsing time}{parsing time vs file size vs
number lines factor}

\newpage
\subsection{Rule Ordering}

@@

    \item [normal]  Hypothetically the most optimal order: rules which
        match most often will be tried first.

@@

\Graphref{Parsing time relative to shuffled} shows the parsing times of
normal and reversed orderings relative to shuffled ordering.  Overall this
optimisation provides a modest but worthwhile reduction in parsing time of
approximately 10\%.  \showgraph{build/graph-hits}{Hits per rule}{rule hits
graph} \showgraph{build/graph-optimal-and-reverse-vs-shuffle}{Parsing time
relative to shuffled ordering}{Parsing time relative to shuffled}

\noindent\textbf{Why are there dips in the graphs?}

XXX MAYBE REDUCE THIS\@?

\label{Why are there dips in the graphs?}

The dips at log files 22 and 62--68 in \graphref{Parsing time relative to
shuffled} correspond to peaks in log file size in \graphref{Parsing time,
file size, and number of lines}, and peaks in \graphref{parsing time vs
file size vs number lines factor}, where a peak means that performance is
better.  The explanation for this took some time to arrive at, but it turns
out to be reasonably simple.  The large log files in question were caused
by a mail forwarding loop, where the distribution of log lines is quite
different to normal, resulting in different performance characteristics.
The vast majority of log lines when a mail loop occurs are from Postfix
components which have a small number of rules associated with them, whereas
in general the majority of log lines are from the Postfix component
(\daemon{smtpd}) with the highest number of rules.  Furthermore in normal
log files the majority of log lines are distributed much more evenly across
rules than the majority of log lines during a mail loop.

\newpage

These two characteristics combine to reduce the average number of rules
required to parse a log line when there is a mail loop, as shown by the
peaks in \graphref{parsing time vs file size vs number lines factor} and
troughs in \graphref{Parsing time relative to shuffled}.  When the rule
ordering is reversed the majority of log lines generated by a mail loop
will be parsed using very few rules, because there are very few rules for
the Postfix components producing the log lines.  Without a mail loop the
Postfix components producing the majority of log lines have a large number
of rules associated with them, and reverse ordering deliberately starts
with the least likely rules and moves towards the most likely.  This leads
to a noticeable drop in the average time required to parse a log line
during a mail loop, as shown in \graphref{Parsing time relative to
shuffled}.  The number of rules which need to be consulted when the
ordering is shuffled varies between the optimal and the pessimal, with the
performance varying proportionally.

@@

How any architecture scales as the number of rules increases is important,
but it is particularly important in this architecture because it is
expected that there will be a large number of rules.  There are
\numberOFrules{} rules in the full Postfix ruleset, whereas the minimum
number of rules required to parse the \numberOFlogFILES{} log files is
\numberOFrulesMINIMUM{}, \numberOFrulesMINIMUMpercentage{} of the total.  A
second set of statistics was generated using the minimum ruleset and the
results compared to the statistics generated using the full ruleset: the
percentage parsing time increase when using the full ruleset instead of the
minimal ruleset for normal, shuffled and reversed orderings is shown in
\graphref{Percentage parsing time increase of maximum ruleset over minimum
ruleset}.
\showgraph{build/graph-full-ruleset-vs-minimum-ruleset}{Percentage parsing
time increase of maximum ruleset over minimum ruleset}{Percentage parsing
time increase of maximum ruleset over minimum ruleset}

It is clear from \graphref{Percentage parsing time increase of maximum
ruleset over minimum ruleset} that the increased number of rules has a
noticeable performance impact with reverse ordering, and to a lesser extent
with shuffled ordering.  The normal ordering (where are the most frequently
matching rules are tried first) shows a mean increase of
\input{build/include-full-ruleset-vs-minimum-ruleset} in parsing time for a
\numberOFrulesMAXIMUMpercentage{} increase in the number of rules.  These
results show that the architecture scales very well as the number of rules
increases, and that sorting the rules is an important optimisation enabling
this.

>>>>

% \showgraph{filename}{caption}{label}
\newcommand{\showgraph}[3]{%
    \begin{figure}[hbtp]%
        \caption{#2}\label{#3}%
        \includegraphics{#1}%
    \end{figure}%
}

%\showtable{filename}{caption}{label}
\newcommand{\showtable}[3]{
    \begin{table}[hbtp]
        \caption{#2}\label{#3}
        \input{#1}
    \end{table}
}

@@

XXX DISCUSS COMPLEXITY\@?

XXX MAYBE MERGE THE AUTHOR SECTIONS IF DESPERATE\@.

@@

Existing parsers were considered, but their parsing coverage was too low,
their parsing too inexact, and/or they did not extract sufficient data.
The effort required to adapt and improve an existing parser was judged to
be greater than the effort to write one from scratch, as the techniques
used by the existing parsers severely limited their potential: some ignored
the majority of log lines, parsing specific log lines with great accuracy,
but in an inextensible fashion; others sloppily parsed the majority of log
lines, but were incapable of distinguishing between log lines of the same
type, e.g.\ rejections, and extracted only a portion of the data required.
The only prior published work on the subject of parsing Postfix log files
is~\cite{log-mail-analyser}, which aims to present correlated log data in a
form suitable for a systems administrator to search using the myriad of
standard Unix text processing utilities already available.  A full state of
the art review is outside the scope of this paper but will be included in
the thesis resulting from this work.

@@

The parser does not try to detect overlapping rules; that responsibility is
left to the author of the rules.  Unintentionally overlapping rules lead to
inconsistent parsing and data extraction because the first matching rule
wins, and the order in which rules are tried against each line may change
between log files.  Overlapping rules are frequently a requirement,
allowing a more specific rule to match some log lines and a more general
rule to match the majority, e.g.\ separating SMTP delivery to specific
sites from SMTP delivery to the rest of the world.  The algorithm provides
a facility for specifying the order of overlapping rules: the priority
field in each rule (defaults to zero).  Rules are sorted by priority,
highest first, and then rules with the same priority are sorted by the
number of successful matches they had parsing the previous log file.
Negative priorities may be useful for catchall rules.

Storing the rules separately to the parser allows external tools to be
written to detect overlapping rules.  Traditional regexes are equivalent in
computational power to Finite Automata (FA); it is possible to convert
between regexes and FA, so regex overlap could be calculated by finding the
intersection of two FA XXX REFERENCE\@.  Perl 5.10 regexes are more
powerful than traditional regexes, e.g.\ it is possible to match correctly
nested and balanced brackets to an arbitrary depth:
\newline{}\tab{}\verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/!\newline{}
Perl 5.10 regexes can maintain an arbitrary state stack and are thus
equivalent in computational power to Pushdown Automata (PA), so overlap
detection would require calculating the intersection of two PA.\@  XXX
REFERENCE OR SOMETHING\@.

XXX OVERLAP DETECTION\@: L1 n L2 = E* --- ((E* --- L1) u (E* --- L2)).
CONTAINMENT CANNOT BE DETERMINED FOR CFGs, INTERSECTION IS POSSIBLE\@.

@@

    \item [Partially context-sensitive actions and cascaded parsing:]
        Actions can consult the results (or lack of results) of previous
        actions when executing, providing some context sensitivity.
        Actions can not inspect past or future inputs, however actions
        can return a modified input line that will be parsed as if read
        from the input stream, allowing for a simplified version of
        cascaded parsing~\cite{cascaded-parsing}.

@@

An important property of a parser is how parsing time scales relative to
input size: does it scale linearly, polynomially, or exponentially?
\Graphref{Parsing time, file size, and number of lines} shows the parsing
time in seconds, file size in MB, and tens of thousands of log lines for
each of the \numberOFlogFILES{} log files.  All three lines run roughly in
parallel, giving the impression that the algorithm scales linearly with
input size.  This impression is borne out by \graphref{parsing time vs file
size vs number lines factor} which plots both the ratio of file size vs
parsing time, and the ratio of number of log lines vs parsing time (higher
is better in both cases).  The ratios are quite tightly banded across the
graph, showing that the algorithm scales linearly.  The ratio increases
(i.e.\ improves) for log files 22 and 62--68 despite their larger than
average size (shown in \graphref{Parsing time, file size, and number of
lines}).  \showgraph{build/graph-input-size-vs-parsing-time}{Parsing time,
file size, and number of lines}{Parsing time, file size, and number of
lines} \showgraph{build/graph-input-size-vs-parsing-time-ratio}{Ratio of
file size and number of lines to parsing time}{parsing time vs file size vs
number lines factor}

\subsection{Rule Ordering}

@@

    \item [optimal]  Hypothetically the most optimal order: rules which
        match most often will be tried first.

@@

\Graphref{Parsing time relative to shuffled} shows the parsing times of
optimal and reversed orderings relative to shuffled ordering.  Overall this
optimisation provides a modest but worthwhile reduction in parsing time of
approximately 10\%.  \showgraph{build/graph-hits}{Hits per rule}{rule hits
graph} \showgraph{build/graph-optimal-and-reverse-vs-shuffle}{Parsing time
relative to shuffled ordering}{Parsing time relative to shuffled}

@@

How any architecture scales as the number of rules increases is important,
but it is particularly important in this architecture because it is
expected that there will be a large number of rules.  There are
\numberOFrules{} rules in the full Postfix ruleset, whereas the minimum
number of rules required to parse the \numberOFlogFILES{} log files is
\numberOFrulesMINIMUM{}, \numberOFrulesMINIMUMpercentage{} of the total.  A
second set of statistics was generated using the minimum ruleset and the
results compared to the statistics generated using the full ruleset: the
percentage parsing time increase when using the full ruleset instead of the
minimal ruleset for optimal, shuffled and reversed orderings is shown in
\graphref{Percentage parsing time increase of maximum ruleset over minimum
ruleset}.
\showgraph{build/graph-full-ruleset-vs-minimum-ruleset}{Percentage parsing
time increase of maximum ruleset over minimum ruleset}{Percentage parsing
time increase of maximum ruleset over minimum ruleset}

It is clear from \graphref{Percentage parsing time increase of maximum
ruleset over minimum ruleset} that the increased number of rules has a
noticeable performance impact with reverse ordering, and to a lesser extent
with shuffled ordering.  The optimal ordering (where the most frequently
matching rules are tried first) shows a mean increase of
\input{build/include-full-ruleset-vs-minimum-ruleset} in parsing time for a
\numberOFrulesMAXIMUMpercentage{} increase in the number of rules.  These
results show that the architecture scales very well as the number of rules
increases, and that sorting the rules is an important optimisation enabling
this.
